Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-18 paper/paper.tex
Finished the sections on the index and the experiments. No proofreading though.

% Queries
\newcommand{\rank}{\ensuremath{\mathsf{rank}}}
\newcommand{\select}{\ensuremath{\mathsf{select}}}
\newcommand{\LF}{\ensuremath{\mathsf{LF}}}
\newcommand{\find}{\ensuremath{\mathsf{find}}}
\newcommand{\locate}{\ensuremath{\mathsf{locate}}}

% Graphs
\newcommand{\gindegree}{\ensuremath{\mathsf{in}}}
\newcommand{\goutdegree}{\ensuremath{\mathsf{out}}}
\newcommand{\glabel}{\ensuremath{\mathsf{label}}}
\newcommand{\gpred}{\ensuremath{\mathsf{pred}}}
\newcommand{\gkey}{\ensuremath{\mathsf{key}}}
\newcommand{\gvalue}{\ensuremath{\mathsf{value}}}
\newcommand{\gnode}{\ensuremath{\mathsf{node}}}

@@

In this paper, we primarily consider sequences over the \emph{DNA} alphabet $\set{\$, \baseA, \baseC, \baseG, \baseT, \baseN}$. Characters $\baseA$, $\baseC$, $\baseG$, and $\baseT$ are called \emph{bases}, while character $\baseN$ represents an arbitrary or unknown base. In addition to the endmarker $\$$, the alphabet may also contain other characters for technical purposes. Each character $c$ of the DNA alphabet has a \emph{complement} $\dnacomp{c}$ defined as $\dnacomp{\baseA} = \baseT$, $\dnacomp{\baseC} = \baseG$, $\dnacomp{\baseG} = \baseC$, $\dnacomp{\baseT} = \baseA$, and $\dnacomp{c} = c$ for other characters $c$. Given a DNA sequence $S$, its \emph{reverse complement} is the sequence $\revcomp{S}$ obtained by reversing the non-technical parts of the sequence and replacing each character with its complement. For example, $\revcomp{\dnaseq{GATTACA}\$} = \dnaseq{TGTAATC}\$$.

@@

The following definitions are useful when discussing path indexes.

\begin{definition}[Prefix-matching strings]
Let $S$ and $S'$ be strings over alphabet $\Sigma$. We say that string $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$.
\end{definition}

\begin{definition}[Prefix-free set]
Let $\mathcal{S}$ be a set of strings over alphabet $\Sigma$. We say that set $\mathcal{S}$ is \emph{prefix-free}, if no two strings $S, S' \in \mathcal{S}$ (with $S \ne S'$) prefix-match.
\end{definition}

@@

The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting arbitrary substring queries. It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges. Although fast and versatile, suffix trees are only of limited use in indexing large texts, as they require much more space than the text itself.

@@

We can use the sampled suffix array pointers to find the text positions containing the occurrences. If $\SA[i]$ is not sampled, we start iterating $\LF(i)$, until we find a sampled pointer. If we find a sample at $\SA[\LF^{k}(i)]$, we know that
$$
\SA[i] = (\SA[\LF^{k}(i)] + k) \bmod n.
$$
If we have sampled one out of $d$ suffix array pointers at regular intervals, finding each occurrence takes $\Oh{d}$ rank queries. If we also sample one out of $d'$ \emph{inverse suffix array} pointers\footnote{The suffix array is a permutation of $\set{0, \dotsc, n-1}$, and the inverse suffix array is the inverse permutation.}, we can \emph{extract} an arbitrary substring $X$ of the text using $\Oh{\abs{X}+d'}$ rank queries. In a typical case, we expect an \FMindex{} to take less than $n \log \sigma$ bits of space and to be able to find about 100,000 pattern occurrences per second \cite{Ferragina2009a}.

@@

Query $G'.\locate(X)$ requires retrieving the values $G'.\gvalue(v')$ for nodes $v' \in V'$ in the lexicographic range returned by query $G'.\find(X)$. Storing the values explicitly for all nodes would make the path index several times larger than the graph structure described above. To avoid that, GCSA uses a similar \emph{sampling} scheme to the one used in FM-indexes. We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are integers, and that the node numbering has been chosen with our sampling scheme in mind. If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.

@@

If $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$, such as in de~Bruijn graphs and maximally pruned de~Bruijn graphs, we can use a simplified encoding for the graph \cite{Siren2014}. We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$. If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set$B_{c}[i] = 1$. Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
$$
G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
$$
We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector. On the other hand, computing $G'.\LF()$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$. The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.

@@

Prefix-doubling; k to 2k, 4k, 8k. original GCSA did not have to store labels, as it built an unlimited order graph

prune, extend, full prune, merge

paths: start, extension, predecessors, key; encode key as a sequence of kmer ranks

prune: only if start node is the same; full prune as before

extend: join when extension matches start; do not extend if already pruned; if right side is pruned, the extended path is also pruned

merge: invariant: path labels are the key set of a pruned de Bruijn graph; full pruning produces a maximally pruned graph

disk-based: each chromosome in a separate file sorted by keys; extend by file; prune/merge essentially merge sorted lists

edges and construction: use \LFmapping{} in GCSA of the de~Bruijn graph of the kmers to determine the rank sequence of $c \cdot G'.\gkey(v')$; need $\sigma + 1$ sequential pointers in each file

@@

Path graphs; LCP array; compact trie of keys; LCP interval tree

count()
- SadaCount \cite{Sadakane2007a}
- compressible \cite{Gagie2015}

parent()
- CST-NPR: \cite{Fischer2009a}
- CST MEMs: \cite{Ohlebusch2010a}
- used for MEM mapping: \cite{Li2013}

@@

@@

% FIXME
\subparagraph*{Acknowledgements.}

I thank Richard Durbin, Erik Garrison, and Adam Novak for \dots

>>>>

% Queries
\newcommand{\rank}{\ensuremath{\mathsf{rank}}}
\newcommand{\select}{\ensuremath{\mathsf{select}}}
\newcommand{\LF}{\ensuremath{\mathsf{LF}}}
\newcommand{\find}{\ensuremath{\mathsf{find}}}
\newcommand{\locate}{\ensuremath{\mathsf{locate}}}
\newcommand{\parent}{\ensuremath{\mathsf{parent}}}
\newcommand{\countq}{\ensuremath{\mathsf{count}}}

% Graphs
\newcommand{\gindegree}{\ensuremath{\mathsf{in}}}
\newcommand{\goutdegree}{\ensuremath{\mathsf{out}}}
\newcommand{\glabel}{\ensuremath{\mathsf{label}}}
\newcommand{\gpred}{\ensuremath{\mathsf{pred}}}
\newcommand{\gkey}{\ensuremath{\mathsf{key}}}
\newcommand{\gvalue}{\ensuremath{\mathsf{value}}}
\newcommand{\gnode}{\ensuremath{\mathsf{node}}}
\newcommand{\gext}{\ensuremath{\mathsf{ext}}}

@@

We primarily consider sequences over the \emph{DNA} alphabet $\set{\$, \baseA, \baseC, \baseG, \baseT, \baseN}$. Characters $\baseA$, $\baseC$, $\baseG$, and $\baseT$ are called \emph{bases}, while character $\baseN$ represents an arbitrary or unknown base. In addition to the endmarker $\$$, the alphabet may also contain other characters for technical purposes. Each character $c$ of the DNA alphabet has a \emph{complement} $\dnacomp{c}$ defined as $\dnacomp{\baseA} = \baseT$, $\dnacomp{\baseC} = \baseG$, $\dnacomp{\baseG} = \baseC$, $\dnacomp{\baseT} = \baseA$, and $\dnacomp{c} = c$ for other characters $c$. Given a DNA sequence $S$, its \emph{reverse complement} is the sequence $\revcomp{S}$ obtained by reversing the non-technical parts of the sequence and replacing each character with its complement. For example, $\revcomp{\dnaseq{GATTACA}\$} = \dnaseq{TGTAATC}\$$.

@@

Let $S$ be a string and $S'$ be be a string or an infinite character sequence over alphabet $\Sigma$. We say that sequences $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$. Set $\mathcal{S}$ of strings is \emph{prefix-free}, if no two strings $S, S' \in \mathcal{S}$ (with $S \ne S'$) prefix-match.

@@

The \emph{suffix tree} \cite{Weiner1973} is the most fundamental full-text index supporting substring queries. It is formed by taking the suffixes of the text, storing them in a trie, and compacting unary paths in the trie into single edges. If $v$ is a node of a suffix tree, we write $\ell(v)$ to denote the label of the path from the root to node $v$. Although fast and versatile, suffix trees are only of limited use in indexing large texts, as they require much more space than the text itself.

@@

We can use the sampled suffix array pointers to find the text positions containing the occurrences. If $\SA[i]$ is not sampled, we start iterating $\LF(i)$, until we find a sampled pointer. If we find a sample at $\SA[\LF^{k}(i)]$, we know that
$$
\SA[i] = (\SA[\LF^{k}(i)] + k) \bmod n.
$$
If we have sampled one out of $d$ suffix array pointers at regular intervals, finding each occurrence takes $\Oh{d}$ rank queries. If we also sample one out of $d'$ \emph{inverse suffix array} pointers\footnote{The suffix array is a permutation of $\set{0, \dotsc, n-1}$, and the inverse suffix array is the inverse permutation.}, we can \emph{extract} an arbitrary substring $X$ of the text using $\Oh{\abs{X}+d'}$ rank queries.

@@

Query $G'.\locate(X)$ requires retrieving the values $G'.\gvalue(v')$ for nodes $v' \in V'$ in the lexicographic range returned by query $G'.\find(X)$ and removing duplicates. Storing the values explicitly for all nodes would make the path index several times larger than the graph structure described above. To avoid that, GCSA uses a similar \emph{sampling} scheme to the one used in FM\nobreakdash-indexes. We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are integers, and that the node numbering has been chosen with our sampling scheme in mind. If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.

@@

If $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$, such as in de~Bruijn graphs and maximally pruned de~Bruijn graphs, we can use a simplified encoding for the graph. We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$. If node $v' \in V$ has a predecessor with label $c \in \Sigma$, we set$B_{c}[i] = 1$. Backward searching is now simpler: $[sp_{i}, ep_{i}] = [G'.\LF(sp_{i+1}, c), G'.\LF(ep_{i+1}+1, c) - 1]$, where
$$
G'.\LF(i, c) = \bvOUT.\rank(C[c] + B_{c}.\rank(i, 1), 1).
$$
We replace two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) with a single $\rank$ query on a bitvector. On the other hand, computing $G'.\LF()$ is now more expensive, as we have to look at $B_{c}[i]$ for all $c \in \Sigma$ to determine $\BWT[i]$. The alternative is to follow outgoing edges when searching for samples; this requires two $\select$ queries per edge.

@@

GCSA construction \cite{Siren2014} is based on the original \emph{prefix-doubling} algorithm for suffix array construction \cite{Manber1993}. The original GCSA started from paths of length $1$ in the input graph, and then repeatedly \emph{joined} paths of length $k$ into paths of length $2k$, until each path had a distinct label. The resulting path graph was essentially an unlimited-order pruned de~Bruijn graph and supported queries of arbitrary length.

We use a variant of that algorithm with GCSA2. Let $G = (V, E)$ be the input graph. We extract all paths of length $k$ (typically for $k = 16$) from graph $G$. For each path $P = v_{0} \dotsm v_{\abs{P}-1}$, we store several fields. \emph{Key} $P.\gkey$ encodes $G.\glabel(P)$ as a sequence of lexicographic ranks of \kmer{k}s. If $\abs{P}$ is not an integer multiple of $k$, the key consists of the \kmer{k} ranks for the lexicographically smallest \kmer{(\lceil \abs{P}/k \rceil \cdot k)} having $G.\glabel(P)$ as a prefix, followed by the rank of the last \kmer{k} in the largest such \kmer{(\lceil \abs{P}/k \rceil \cdot k)}. \emph{Value} $P.\gvalue$ is the start node $v_{0}$ of the path. We store the set of \emph{predecessor labels} $\set{c \in \Sigma \mid \abs{G.\gpred(v_{0}, c)} > 0}$ as $P.\gpred$. For each possible \emph{extension node} $v \in \set{v \in V \mid (v_{\abs{P}-1}, v) \in E}$, we create a new copy of the path and store the node as $P.\gext = v$.

The construction uses several supporting structures. We build an \orderk{k} \emph{de~Bruijn graph} $G_{d} = (V_{d}, E_{d})$ of the path labels and encode it as a GCSA, using the predecessor labels $P.\gpred$ for determining the edges. Let $v_{0}, \dotsc, v_{\abs{V_{d}}-1}$ be the nodes of the de~Bruijn graph in lexicographic order by the keys. We use two additional arrays: \emph{the LCP array} $\LCP[0, \abs{V_{d}}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G_{d}.\gkey(v_{i-1})$ and $G_{d}.\gkey(v_{i})$ (with $\LCP[0] = 0$), and the \emph{last character array} $L[0, \abs{V_{d}}-1]$, where $L[i] = G_{d}.\gkey(v_{i})[k-1]$. The LCP array is stored in a wavelet tree for fast \emph{range minimum queries} \cite{Gagie2012a}.

Because we have to store path labels explicitly in the keys, we only do a limited number of \emph{doubling steps}, typically $2$ or $3$. After $d$ doubling steps, the length of the paths is $2^{d} k$, and we can use them to build a maximally pruned \orderk{(2^{d} k)} de~Bruijn graph. Each doubling step consists of a \emph{pruning} step, followed by an \emph{extension} step. The pruning step uses a limited form of Lemma~\ref{lemma:dbg-prune} and applies it to all possible ranges of paths. Given two paths $P$ and $P'$, we can determine the longest common prefix of the path labels by using the keys $P.\gkey$ and $P'.\gkey$ and the LCP array. If all paths with a certain prefix of their labels start from the same node, we merge them into a single path $Q$ with $Q.\gkey$ based on the shared prefix and $Q.\gext = -1$.

The extension step transforms the current set of paths of length (up to) $k'$ and transforms it into a set of paths of length (up to) $2k'$. If $P$ is a path with $P.\gext = -1$, we use it as such. If $P.\gext = P'.\gvalue$ for paths $P$ and $P'$, we create a new path $PP'$. We set $(PP').\gkey$ according to the concatenation of the path labels, take $\gvalue$ and $\gpred$ from path $P$, and take $\gext$ from path $P'$. If we have another path $Q$ with $Q.\gkey = P.\gkey$ such that $QP'$ is a path, and if $P'.\gext = -1$, all possible \kmer{2k'} extensions of label $G.\glabel(PP')$ are also labels of paths starting from node $Q.\gvalue$, and the other way around. Hence paths $PP'$ and $QP'$ can be represented by a single node in a pruned de~Bruijn graph.

The doubling steps are followed by the \emph{merging steps}, which transforms the paths of length (up to) $k'$ into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$. First we merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$. If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$. We also store the union of predecessor labels as $G'.\gpred(v') = \bigcup_{i=0}^{m-1} P_{i}.\gpred$. Then we apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.

Storing the paths and graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs. To avoid that, we store them on \emph{disk} when possible. Each chromosome forms a subgraph that is disjoint from the other chromosomes, except near the source node and the sink node. Hence we can store the paths corresponding to each chromosome in a separate file, sorted by their labels in lexicographic order. A doubling step can be done separately for each of the chromosomes. A pruning step takes a set of sorted lists paths and merges the lists. It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged. The merged path is written into the new file for that chromosome, and the original range of paths is removed from the buffer. The merging step is mostly the same as a pruning step, but the output format is different.

After storing the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$ on disk, we proceed to \emph{building the index}. Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$ we have stored, but the outdegree sequence $\bvOUT$ requires further processing. Because graph $G''$ is a path graph, there is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in G''.\gpred(v')$. We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers. The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a $c \in \Sigma$. We can also sample the nodes for $\locate$ queries during the same scans.

Checking whether key $G''.\gkey(u'')$ prefix-matches string $c \cdot G''.\gkey(v'')$ can be done by using the GCSA for de~Bruijn graph $G_{d} = (V, E)$ and the last character array $L$. If the lexicographic rank of \kmer{k} $X$ is $i$, the lexicographic rank of \kmer{k} $(cX)[0, k-1]$ is $G_{d}.\LF(i, c)$. If \kmer{2k} $X$ is encoded with \kmer{k} ranks $(i, j)$, we can encode string $cX$ as a lexicographic range of \kmer{k} rank sequences, with $(G_{d}.\LF(i, c), G_{d}.\LF(j, L[i]), G_{d}.\LF(0, L[j]))$ as the lower bound and $G_{d}.\LF(\abs{V_{d}}-1, L[j])$ as the last rank in the upper bound. String $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ if and only the corresponding lexicographic ranges of \kmer{k} rank sequences overlap.

@@

GCSA is a generalization of FM\nobreakdash-indexes for path graphs. Because path graphs are quite similar to strings, at least when compared with other labeled graphs, we can often generalize algorithms using FM\nobreakdash-indexes and data structures extending them to work with path graphs.

Let $G' = (V', E')$ be an \orderk{k} (pruned) de~Bruijn graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order by their keys. We define the \emph{LCP array} of graph $G'$ in the same way as in GCSA construction: $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. The following extensions are based on the trie of keys $G'.\gkey(v')$ for all nodes $v' \in V'$, with unary paths compacted in the same way as in the suffix tree. The compacted trie can be considered the \emph{suffix tree of the path graph}.

The \emph{LCP interval tree} \cite{Abouelhoda2004} is an alternate representation of the suffix tree. Each node $v$ of the suffix tree is represented by the lexicographic range $[sp, ep]$ matching the label $\ell(v)$. These ranges can be determined from the LCP array. If we augment the FM\nobreakdash-index with an LCP array supporting \emph{next/previous smaller value} queries and \emph{range minimum queries}, we get a space-efficient data structure providing the full functionality of the suffix tree \cite{Fischer2009a}.

% FIXME justify why we can use parent queries
% we need a lemma stating that we always match entire equivalence classes
If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range of the LCP array, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. However, if pattern $X$ of length $\abs{X} \le k$ matches lexicographic range $[sp, ep]$, we can safely use a $\parent$ query for the corresponding node in the suffix tree.

BWA-MEM \cite{Li2013} aligns reads to reference genomes by finding \emph{maximal exact matches} between the read and the reference. It stores both the reference and its reverse complement are stored in the same index, which allows it to extend the pattern in both directions. GCSA cannot use similar techniques, as we cannot guarantee that the length of the lexicographic range matching pattern $X$ is the same as the length of the range matching its reverse complement $\revcomp{X}$ (that the key set contains key $K$ if and only if it contains the reverse complement $\revcomp{K}$). Instead of using bidirectional BWT, we search for maximal exact matches using \LFmapping{} and the $\parent$ operation \cite{Ohlebusch2010a}. We have implemented a GCSA-based read aligner using that algoritm in vg \cite{Garrison2014-2016}.

In an ordinary FM\nobreakdash-index, the length of a lexicographic range tells the number of distinct pointers in the range. In GCSA, each node $v' \in V'$ may have multiple values (pointers), and each value may occur in multiple nodes. Determining the  number of distinct start nodes with a $\locate$ query can be slow. Hence we need additional structures to support efficient \emph{counting queries}. The problem is similar to determining the \emph{document frequency} of a pattern. Given a collection of documents, the document frequency of pattern $X$ is the number of distinct documents that contain occurrences of the pattern. We can augment an FM\nobreakdash-index with a bitvector of length $2n-d$, where $n$ is the total size of the document collection and $d$ is the number of documents, to compute document frequencies efficiently \cite{Sadakane2007a}.

Let $v$ be a suffix tree node corresponding to lexicographic range $[sp, ep]$ in the collection, and let $\countq(v)$ be the document frequency of the label $\ell(v)$. If nodes $v_{0}, \dotsc, v_{m-1}$ are the children of node $v$, we determine the number of \emph{redundant} documents in node $v$ as $R(v) = \sum_{i=0}^{m-1} \countq(v_{i}) - \countq(v)$. Consider now the \emph{inorder traversal} of the suffix tree, where we may encounter the same nodes multiple times. We create an array $R[0, n-2]$ of the values $R(v)$ over all internal nodes $v$. If the $i$th internal node we encounter is our first visit to node $v$, we set $R[i] = R(v)$. For subsequent visits to the same node, we set $R[j] = 0$. Range $R[sp, ep-1]$ now covers the internal nodes in the subtree rooted at node $v$. We can now determine document frequencies as $\countq(v) = (ep+1-sp) - \sum_{i=sp}^{ep-1} R[i]$. If we encode array $R$ in unary, with value $x$ becoming $0^{x} 1$, we get a bitvector $B_{R}[0, n-d-1]$, where we can compute the sums with $\select$ queries:
$$
\sum_{i=a}^{b} R[i] = (B_{R}.\select(b+1, 1) - b) - (B_{R}.\select(a, 1) + 1 - a).
$$

For value counting, we use array $R$ to store the number of redundant values in each suffix tree node. We also use another array $A[0, \abs{V'}-1]$, which tells the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$. We then encode array $A$ as a bitvector $B_{A}$ in the same way as bitvector $B_{R}$ above. The number of distinct start nodes for paths matching pattern $X$ (with $\abs{X} \le k$) is then
$$
G'.\countq(X) = G'.\countq(G'.\find(X)) = (ep+1-sp) + \sum_{i=sp}^{ep} A[i] - \sum_{i=sp}^{ep-1} R[i],
$$
where $[sp, ep]$ is the lexicographic range returned by the $\find$ query. The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.

@@

GCSA2 has been implemented as the path index library of vg \cite{Garrison2014-2016}. The implementation is in C++ and the source code is available on GitHub (\url{https://github.com/jltsiren/gcsa2}). As the implementation uses the \emph{Succinct Data Structures Library 2.0} (SDSL) \cite{Gog2014b} and \emph{libstdc++ parallel mode}, it requires g++ version~4.9 or later to compile. As a general design principle, fast succinct data structures have been chosen over the smaller but slower compressed data structures.

The \emph{variation} graphs used in vg have strings as node labels. The average length of a node label is typically around 10~bases. Each node can be traversed both in forward orientation and in reverse complement orientation, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around. When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels. Input graph nodes corresponding to first offsets of variation graph nodes are always selected as samples for performance reasons.

The following experiments were done on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256~gigabytes of memory running Ubuntu~12.04 on Linux kernel~3.2.0. Input files, final indexes, and temporary files were all stored on a distributed Lustre file system. We used vg version~1.3.0 for processing the graphs, GCSA2 version~0.6 for building the indexes, and GCSA2 version~0.6.1 for the query benchmarks.

We built vg graphs based from the human reference genome (GRCh37) and the 1000~Genomes Project phase~3 variation \cite{1000GP2015}. To avoid excessive growth, we pruned complex regions of the graphs using \texttt{vg mod -p -l 16 -e 4 | vg mod -S -l 100}, removing paths where \kmer{16}s cross more than $4$ nontrivial edges and subgraphs shorter than $100$ bases. We then extracted paths of length $16$ from the forward strands of all chromosomes. There were a total of 4.80~billion paths with 1.53~billion distinct labels. We then built GCSA2 using $1$, $2$, and $3$ doubling steps, producing \orderk{32}, \orderk{64}, and \orderk{128} indexes, respectively. The results are summarized in Table~\ref{table:construction}.

\begin{table}[t!]
\begin{center}
\caption{GCSA2 construction with different orders of the path graph. Billions of nodes in the de~Bruijn graph and the pruned de~Bruijn graph; index size in gigabytes; construction time in hours; peak memory and disk usage in gigabytes; and disk I/O volume for reading and writing in terabytes.}\label{table:construction}
\begin{tabular}{c|ccc|ccc|cc}
\hline
\textbf{Order} & \textbf{Nodes} & \textbf{Pruned} & \textbf{Index} & \textbf{Time} & \textbf{Memory} & \textbf{Disk} & \textbf{Read} & \textbf{Write} \\
\hline
 32 & 6.23G & 4.39G & 13.7 GB & 9.85 h & 63.2 GB & 401 GB & 1.55 TB & 1.05 TB \\
 64 & 16.9G & 5.27G & 14.2 GB & 13.6 h & 56.1 GB & 424 GB & 2.23 TB & 1.71 TB \\
128 &  118G & 5.76G & 15.3 GB & 18.0 h & 56.9 GB & 489 GB & 3.00 TB & 2.47 TB \\
\hline
\end{tabular}
\end{center}
\end{table}

A single-strand whole-genome index can be built overnight with 96~gigabytes of memory. Even 64~gigabytes is enough, but the lack of sufficient disk cache may make the construction slower. The size of the de~Bruijn graph grows exponentially at roughly $1.031^{k} \cdot 2.38$~billion nodes, while the growth rate of the pruned de~Bruijn graph is much slower. The index itself grows even more slowly with $k$, as we sample the values more often in lower order graphs.

We extracted \kmer{16}s and \kmer{32}s from the vg graphs for benchmarking purposes using \texttt{vg sim}. We then queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned during construction. The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index. The main reason is that we use to use $\select$ queries, which are much slower than $\rank$ queries on plain bitvectors \cite{Gog2014}.

\begin{table}[t!]
\begin{center}
\caption{Performance benchmarks using an \orderk{64} GCSA2. Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in milliseconds.}\label{table:benchmark}
\begin{tabular}{c|cc|cc|cccc}
\hline
\textbf{Length} & \textbf{Patterns} & \textbf{Matches} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
16 & 351584 & 347453 & 1830M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 0.47 ms \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

\subparagraph*{Acknowledgements.}

I thank Erik Garrison, Richard Durbin, and Adam Novak for the fruitful discussions we had while I developed the GCSA2 index. Many extensions over the original GCSA and clarifications of the assumptions I made are a direct result of those discussions.
